Journey: A Study in Unhelpful Titles (TODO: should we get a new title?)

### 2017-04-21 paper.tex

\title{Journey: A Study of Studying Large Memory Systems (TODO: should we get a new title?)}
\author{Mark Mansi, Suhas Pai, Hasnain Ali Pirzada}
\date{}

@@

@@

@@

We run several different microbenchmarks, each of which has a different memory
allocation pattern. These benchmarks are designed to stress the memory
management system in different ways, rather than simulate real workloads. For
each workload, we measure the latency and increase in memory usage due to each
operation in the benchmark. Notably, we do not touch the pages that we allocate,
avoiding both a page fault and the allocation of a back physical page for the
allocated virtual pages. This allows us to measure just the overhead of memory
management, as opposed to physical page allocation, swapping, page faults, or
other overheads.  Each workload contains $2^{20}$ memory management operations.

The specifications of the machine are listed in Figure \ref{fig:specs}. We
disabled Intel SpeedStep and TurboBoost. All experiments are pinned to the same
core on our machine. Each benchmark was run 5 times and the results aggregated
as described below. We reboot the machine when switching to a different
benchmark, but keep the machine running continually during the runs of the same
benchmark. Each run is executed in isolation, without any additional processes
running beyond services that start when the system is booted and a single ssh
session. We are careful to avoid opening extra ssh sessions, running screen or
tmux, or executing extra commands while a benchmark is running, as these may
impact memory usage, and thus our results.

We measured latency using the \texttt{rdtsc} instruction provided by x86\_64
processors, which gives high resolution cycle-level timestamps. We take a
timestamp before and after each memory management operation in the workload and
take the difference to get latency. The median latency of each operation is
reported across all 5 runs.

To measure memory, we use the procfs's reporting on current memory usage, along
with the slab allocator's count of active \texttt{vm\_area\_struct}s in the
system. Measuring kernel memory usage and associating memory usage with
processes is very difficult. Our methodology assumes that our benchmarks running
in otherwise idle systems should dominate increases in memory usage.  Thus, we
compute total memory usage in the system as the total amount of memory less the
amount of memory used. From this amount, we subtract the resident set size of
the benchmark process. The result is the memory usage of the rest of the system
$R$. We denote $R_i$ as the $i$-th measurement of $R$ during a run ($i = 0, 1,
... 2^{20}$). System memory usage is known to jitter in Linux (TODO: cite). To
adjust for this jitter, we define $\Delta_i = R_i - R_0$. For each operation, we
take the median value of $\Delta_i$ across the 5 runs to reduce jitter further.
This is the value graphed in the figures in this section. $\Delta_i$ represents
the increase in memory usage due to memory management overheads caused by
running $i$ memory management operations.

@@

In this experiment, a single process touches 32GB of virtual memory, leading to
the actual allocation of physical memory. Notice that 32GB is twice as large as
the 16GB of physical memory in the test machine, causing Linux to start paging
after physical memory is exhausted. Also, notice that the Linux kernel will
reserve some memory for its own operation.

@@

This experiment is repeated with 5 different swapfile sizes to capture the
impact of swapfile size on swapping overhead.

(TODO: strided access vs random access?)

@@

>>>>

\title{Journey: A Study in Unhelpful Titles (TODO: should we get a new title?)}
\author{Mark Mansi, Suhas Pai, Hasnain Ali Pirzada}
\date{}

@@

(TODO: explain struct page?)

@@

The specifications of the machine we used to run our experiments are listed in
Figure \ref{fig:specs}. We disabled Intel SpeedStep and TurboBoost and ensure
that the processor frequency stays constant (TODO: did we do this?). All
experiments are pinned to the same core on our machine.

We run several different microbenchmarks, each of which has a different memory
allocation pattern. These benchmarks are designed to stress the memory
management system in different ways, rather than simulate real workloads. 
Each run is executed in isolation, without any additional processes
running beyond services that start when the system is booted and a single ssh
session. We are careful to avoid opening extra ssh sessions, running screen or
tmux, or executing extra commands while a benchmark is running, as these may
impact memory usage, and thus our results.

We measured latency using the \texttt{rdtsc} instruction provided by x86\_64
processors, which gives high resolution cycle-level timestamps. We take a
timestamp before and after each memory management operation in the workload and
take the difference to get latency. (TODO: is this still true?)

To measure memory usage, we use the procfs's reporting on current memory usage.
Where appropriate, we record the slab allocator's count of active
\texttt{vm\_area\_struct}s in the system. Measuring kernel memory usage and
associating memory usage with processes is difficult. Our methodology assumes
that our benchmarks are running in an otherwise idle system and should dominate
increases in memory usage.  We compute total memory usage by the rest of the
system as the total amount of memory less the amount of memory used by the
benchmark.

Finally, we keep transparent huge pages on throughout our experiments, unless
otherwise specified.

@@

For each workload, we measure the latency and increase in memory usage due to
each operation in the benchmark. Notably, we do not touch the pages that we
allocate, avoiding both a page fault and the allocation of a back physical page
for the allocated virtual pages. This allows us to measure just the overhead of
memory management, as opposed to physical page allocation, swapping, page
faults, or other overheads.  Each workload contains $2^{20}$ memory management
operations.

Each benchmark was run 5 times and the results aggregated as described below. We
reboot the machine when switching to a different benchmark, but keep the machine
running continually during the runs of the same benchmark.

Assume the memory usage of the rest of the system is $R$. We denote $R_i$ as the
$i$-th measurement of $R$ during a run ($i = 0, 1, ... 2^{20}$). System memory
usage is known to jitter in Linux (TODO: cite). To adjust for this jitter, we
define $\Delta_i = R_i - R_0$. For each operation, we take the median value of
$\Delta_i$ across the 5 runs to reduce jitter further.  This is the value
graphed in the figures in this section. $\Delta_i$ represents the increase in
memory usage due to memory management overheads caused by running $i$ memory
management operations.

@@

A single process allocates 18GB of virtual memory space. It then proceeds to
write one byte of each 4KB page in the region, leading to the actual allocation
of physical memory. Notice that 18GB exceeds the 16GB of physical memory in the
test machine, causing Linux to start paging after physical memory is exhausted.
Also, notice that the Linux kernel will reserve some memory for its own
operation.

We run this experiment with both a strided access pattern and a random access
pattern. The stride access pattern simply touches pages in order of increasing
address. The random access pattern chooses a random address. Both patterns touch
the same number of addresses.

@@

Moreover, we measure the latency of each memory store operation to see the
impact of swapping and virtual memory usage on memory access times.

We repeat this with 32GB of swap space and 128GB of swap space to see the
impact of swap space size on the amount of memory the kernel reserves for
itself, along with the impact on latency.

@@

These experiments are designed to measure the processing overhead of kswapd when
the system is under high memory pressure.

A single process allocates 18GB of virtual memory space. It then proceeds to
write one byte of each 4KB page in the region, leading to the actual allocation
of physical memory. The process chooses addresses randomly, as in the previous
set of experiments, inducing the maximum amount of swapping.

In all these experiments, the system is given 32GB of swap space total. We
measure the impact of physical memory size and transparent huge pages. We run
four experiments:

\begin{enumerate}
    \item 1GB physical memory with huge pages enabled
    \item 16GB physical memory with huge pages enabled
    \item 1GB physical memory with huge pages disabled
    \item 16GB physical memory with huge pages disabled
\end{enumerate}

To run the 1GB memory benchmarks, we hot unplug 15GB using the sysfs in Linux.
This makes the unplugged memory unusable by the whole system (kernel and user).
